The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (9th edition -- distributed with CrypTool version 1.4.20)} \vspace{60pt}

### 2008-07-03 script/en/chapters/numbertheory.tex
1.4.20 updates

Prime numbers can be factorised in a unique (``{\em trivial}'') way:
\begin{eqnarray*}
5 & = & 1 * 5 \nonumber \\
17 & =  & 1 * 17 \nonumber \\
1,013 &  = & 1 * 1,013 \nonumber \\
1,296,409 & = & 1 * 1,296,409. \nonumber
\end{eqnarray*}

@@

Examples of the decomposition of such numbers into prime 
factors\index{Prime factor!decomposition}:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1,029 & = & 3 * 7^3  \nonumber \\
5,324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

\newpage
{\bf Examples:}
\begin{itemize}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{N} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{N} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{itemize}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

{\bf Example: }\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 \equiv 9 \equiv 0$ (mod $3$); but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem \ref{thm-zth-div} does not even apply to sums in one direction. It is correct for sums only if the 
remainder is 0 and only in the following direction:  if a divisor divides both summands with no remainder, it also 
divides the sum with no remainder.

@@

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$
\end{itemize}
The order in which order the modulo operation is performed is irrelevant.
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The property of closeness is always defined in relation to an operation in a set. See 
\hyperlink{Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

{\bf Example 1:}
$$ (4^3)^2 = 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
\begin{quote}
(1) We can speed up\footnote{The time required to calculate the multiplication 
of two numbers normally depends on the length of the numbers.  We can observe
this if we use the school method to calculate, for instance, $474*228$.  The 
time required increases 
%quadratically
in a quadratic square manner , because we need to multiply $3*3$ numbers. The 
numbers become considerably smaller if we reduce the interim result.} the 
calculation by reducing the {\bf interim results} modulo $5$ but we need 
to take care because not everything will then work in the same way as in 
standard arithmetic.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2\;{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

(3) Therefore, the rule is slightly different for consecutive powers in modular arithmetic: we do not multiply the 
exponents in (mod $m$) but rather in (mod $J(m)$).

@@

Example for $a = 3$ modulo $10$ with $\mathbb{Z}_n^* = \{ 1, 3, 7, 9 \}$ :
\begin{eqnarray*}
3 & \equiv & 3 * 1{\rm ~(mod~}10), \nonumber \\
9 & \equiv & 3 * 3{\rm ~(mod~}10), \nonumber \\
1 & \equiv & 3 * 7{\rm ~(mod~}10), \nonumber \\
7 & \equiv & 3 * 9{\rm ~(mod~}10). \nonumber 
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

Mathematicians often ask, in which conditions the repeated application of an operation results in the neutral
element (compare patterns and structures before)\index{Structure}.

For the $i$-times successive modular multiplication of a number $a$ with $i=1,\cdots, m-1$ the product is the neutral element
of the multiplication (1) if and only if $a$ and $m$ are relatively prime. The value of $i$, for which  
the product $a^{i}=1,$ is called the multiplicative order of $a$. \index{Order!multiplicative}

@@

\begin{definition}\label{def-zth-ordn}
The {\bf multiplicative order} ${\rm ord}_m(a)$ of a whole number $a$ (mod $m$)
(where $a$ and $m$ are co-prime\index{Number!co-prime}) is the 
smallest whole number $e$ for which $a^{e} \equiv 1 ~(mod~m)$.  
\end{definition}

@@

\subsubsection*{Table 4: Values of $a^i {\rm ~mod~} 11,  1 \leq a,i<11$ and according order of $a$ mod $m$:}\label{SrcArith3a}
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
              & i=1 & i=2 & i=3 & i=4 & i=5 & i=6 & i=7 & i=8 & i=9 & i=10  & $ord_{11}(a)$\\
\hline
\hline
$a=1$           & 1  & 1    & 1  & 1    & 1    & 1    & 1  & 1    & 1  & 1     & 1   \\
\hline
$a=2$           & 2  & 4    & 8  & 5   & 10    & 9    & 7  & 3    & 6  & 1    & 10  \\
\hline
$a=3$           & 3  & 9    & 5  & 4 & {\bf 1} & 3    & 9  & 5    & 4  & 1     & 5   \\
\hline
$a=4$           & 4  & 5    & 9  & 3 & {\bf 1} & 4    & 5  & 9    & 3  & 1    & 5 \\
\hline
$a=5$           & 5  & 3    & 4  & 9 & {\bf 1} & 5    & 3  & 4    & 9  & 1    & 5   \\
\hline
$a=6$           & 6  & 3    & 7  & 9   & 10    & 5    & 8  & 4    & 2  & 1    & 10  \\
\hline
$a=7$           & 7  & 5    & 2  & 3   & 10    & 4    & 6  & 9    & 8  & 1    & 10  \\
\hline
$a=8$           & 8  & 9    & 6  & 4   & 10    & 3    & 2  & 5    & 7  & 1    & 10  \\
\hline
$a=9$           & 9  & 4    & 3  & 5 & {\bf 1} & 9    & 4  & 3    & 5  & 1    & 5   \\
\hline
$a=10$         & 10  & 1   & 10  & 1   & 10    & 1   & 10  & 1   & 10  & 1    & 2   \\
\hline
\end{tabular}
\end{center}

@@

Since $45$ is not a prime, there is no ``multiplicative order'' for all values of $a$ (e.~g.\ for the numbers that are not 
relatively prime to $45: 3, 5, 6, 9, 10, 12, \cdots,$ because $45 = 3^2*5$).

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\subsection{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA'' $-$ the following remark is meant with humor to show
that each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
%(\hyperlink{appendix-movies}
%{see appendix A3: Movies and Literature with Relation to Cryptography}).
(see appendix \ref{s:appendix-movies}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

@@

We now combine the conjectures from (a) and (b) for $n=p*q$ to show that:
$$ M^{e * d} \equiv M{\rm ~(mod~}n) {\rm ~for~all~} M < n. $$
From (a) and (b) we have $(M^{e * d} - M) \equiv 0 {\rm ~(mod~} p)$ and  $(M^{e * d} - M) \equiv 0 {\rm ~(mod~} q)$.
Therefore, $p$ and $q$ are both factors of the same number $z = (M^{e * d} - M)$.
Since $p$ and $q$ are {\bf distinct} prime numbers, their product must also be a factor of this number $z$. Thus:
$$
(M^{e * d} - M) \equiv 0 {\rm ~(mod~}p*q) {\rm ~~or~~ } M^{e * d} \equiv M {\rm ~(mod~}p*q) {\rm ~~or~~} 
 M^{e * d} \equiv M {\rm ~(mod~}n).
$$
\hfill$\Box$

@@

$O(l)$ is the average number of processor steps depending on the bit length
$l$ of the number $n$ to be factorised.  For the best currently known
factorisation algorithm the constant $c = (64/9)^{1/173} = 1,923$.

@@

The task of the researchers was not initiated by a challenge, but
they wanted to find the last prime factors of the Mersenne number $2^{1039}+1$
(see ``Wanted List'' of the Cunningham Project\index{%
Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}\\
Cunningham table: \href{http://homes.cerias.purdue.edu/~ssw/cun/pmain1206}
                {\texttt{http://homes.cerias.purdue.edu/\~{}ssw/cun/pmain1206}}\\
The numbers in the Cunningham table have the following syntax:\\
``(2,n)-'' means $2^{n}-1$;~~~
``(2,n)+'' means $2^{n}+1$.\\
To describe the magnitude one writes $p<n>$ or $c<n>$: 
``n'' is the number of decimal digits and ``p'' and ``c'' tell,
whether the number is prime or composite.\\
$2^{1039}-1 = p7 * c307 = p7 * p80 * p227$ \\
It is explained more precisely at the CUN page:\\
``2,651+ means $2^{651} + 1$ and the size (c209 means 209 decimal digits)
of the
number which was factored.  Then come the new factor(s), the discoverer and
the method used.  Recently, only the multiple polynomial quadratic sieve
(ppmpqs), the elliptic curve method (ecm) and the number field sieve (nfs)
have been used.  `hmpqs' stands for hypercube multiple polynomial quadratic
sieve.  Under `new factors', `p90' means a 90-digit prime and `c201' is a
201-digit composite number.''.
}).\\

@@

{\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3,713; ~J(n) = (p-1)*(q-1)=3,588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3,587$ and must be relatively prime
                to $3,588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3,589 \equiv 1{\rm ~(mod~}3,588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{Appendix_A}{appendix A of this chapter}}. \\

@@

This $28$-digit number is divided into $4$-digit parts
(because $2,626$ is still smaller than $n=3,713$):\\
{\tt 0120 2001 0311 0001 2000 0401 2314}

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3,713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code to do RSA
encryption using Mathematica and Pari-GP. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

{\bf 5. Decryption}: \\
Cipher text: {\tt 1404 2932 3536 0001 3284 2280 2235 }\\
This $28$-digit number is divided into $4$-digit parts.\\
All 7 parts are decrypted using:  $M \equiv C^{97}{\rm ~(mod~}3,713)$: \\
{\tt 0120 2001 0311 0001 2000 0401 2314} \\
The 2-digit numbers are transformed into capital letters and blanks.

Using the selected values it is easy for a cryptanalyst\index{Cryptanalysis}
to derive the secret values from the public 
parameters $n=3,713$ and $e=37$ by revealing that $3,713 = 47 * 79$.\\  
If $n$ is a $768$-bit number, there is, according to present knowledge, little chance of this.

@@

Blocks are formed\footnote{\vskip +3 pt \tt \begin{tabular}{lll}
& binary representation  & decimal representation\\
01010010, 82 & 01010010 01010011 & =21075 \\
01010011, 83 & \\
01000001, 65 & 01000001 00100000 & =16672  \\
00100000, 32  \\
01110111, 119 & 01110111 01101111 & =30575 \\
01101111, 111 \\ 
01110010, 114 & 01110010 01101011 & =29291 \\
01101011, 107 \\
01110011, 115 & 01110011 00100001 & =29473 \\
00100001, 33: 
\end{tabular}} (each ASCII character is encoded into a 8 digit binary number below):\\
{\tt 21075 16672 30575 29291 29473}\footnote{Using CrypTool\index{CrypTool} 
you can solve this with the menu
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration} 
with the following options: all 256 ASCII characters, b-adic, block length\index{Block length} 2 and decimal representation.}

@@

Each block is decrypted using: $M \equiv C^{2,473}{\rm ~(mod~}67,519)$: \\
{\tt 82083 65032 119111 114107 115033}

>>>>

Prime numbers can be factorised in a unique (``{\em trivial}'') way:
\begin{eqnarray*}
5 & = & 1 * 5 \nonumber \\
17 & =  & 1 * 17 \nonumber \\
1013 &  = & 1 * 1013 \nonumber \\
1,296,409 & = & 1 * 1,296,409. \nonumber
\end{eqnarray*}

@@

Examples of the decomposition of such numbers into prime 
factors\index{Prime factor!decomposition}:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1029 & = & 3 * 7^3  \nonumber \\
5324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

\newpage
{\bf Examples:}
\begin{itemize}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{itemize}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

{\bf Example: }\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$); but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem \ref{thm-zth-div} does not even apply to sums in one direction. It is correct for sums only if the 
remainder is 0 and only in the following direction:  if a divisor divides both summands with no remainder, it also 
divides the sum with no remainder.

@@

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which order the modulo operation is performed is irrelevant.
\end{itemize}
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The property of closeness is always defined in relation to an operation in a set. See 
\hyperlink{Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

{\bf Example 1:}
$$ (4^3)^2 = 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
\begin{quote}
(1) We can speed up\footnote{The time required to calculate the multiplication 
of two numbers normally depends on the length of the numbers.  We can observe
this if we use the school method to calculate, for instance, $474*228$.  The 
time required increases 
%quadratically
in a quadratic square manner , because we need to multiply $3*3$ numbers. The 
numbers become considerably smaller if we reduce the interim result.} the 
calculation by reducing the {\bf interim results} modulo $5$ but we need 
to take care because \textit{not} everything will then work in the same way as in 
standard arithmetic.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2\;{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

(3) Therefore, the rule is slightly different for consecutive powers in modular arithmetic: We do not multiply the 
exponents in (mod $m$) but rather in (mod $J(m)$).

@@

Example: For $a = 3$ modulo $10$ with $\mathbb{Z}_n^* = \{ 1, 3, 7, 9 \}$ it holds that $a^{-1} = 7$:
\begin{eqnarray*}
3 & \equiv & 3 * 1{\rm ~(mod~}10), \nonumber \\
9 & \equiv & 3 * 3{\rm ~(mod~}10), \nonumber \\
1 & \equiv & 3 * 7{\rm ~(mod~}10), \nonumber \\
7 & \equiv & 3 * 9{\rm ~(mod~}10). \nonumber 
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\hypertarget{patternsandstructures}{}
\subsubsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

@@

Mathematicians often ask, in which conditions the repeated application of an operation results in the neutral
element (compare \hyperlink{patternsandstructures}{Patterns and Structures}, chapter \ref{patternsandstructures})\index{Structure}.

For the $i$-times successive modular multiplication of a number $a$ with $i=1,\cdots, m-1$ the product is the neutral element
of the multiplication (1) if and only if $a$ and $m$ are relatively prime.

\begin{definition}\label{def-zth-ordn}
The {\bf multiplicative order} ${\rm ord}_m(a)$ of a whole number $a$ (mod $m$)
(where $a$ and $m$ are co-prime\index{Number!co-prime}) is the 
smallest whole number $i$ for which $a^{i} \equiv 1 ~(mod~m)$.  
\end{definition}

@@

\subsubsection*{Table 4: Values of $a^i {\rm ~mod~} 11,  1 \leq a,i<11$ and according order of $a$ mod $m$:}\label{SrcArith3a}
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
              & i=1 & i=2 & i=3 & i=4 & i=5 & i=6 & i=7 & i=8 & i=9 & i=10  & $ord_{11}(a)$\\
\hline
\hline
a=1           & {\bf 1}  & 1    & 1  & 1    & 1    & 1    & 1  & 1    & 1  & 1     & 1   \\
\hline
a=2           & 2  & 4    & 8  & 5   & 10    & 9    & 7  & 3    & 6  & {\bf 1}    & 10  \\
\hline
a=3           & 3  & 9    & 5  & 4 & {\bf 1} & 3    & 9  & 5    & 4  & 1     & 5   \\
\hline
a=4           & 4  & 5    & 9  & 3 & {\bf 1} & 4    & 5  & 9    & 3  & 1    & 5 \\
\hline
a=5           & 5  & 3    & 4  & 9 & {\bf 1} & 5    & 3  & 4    & 9  & 1    & 5   \\
\hline
a=6           & 6  & 3    & 7  & 9   & 10    & 5    & 8  & 4    & 2  & {\bf 1}    & 10  \\
\hline
a=7           & 7  & 5    & 2  & 3   & 10    & 4    & 6  & 9    & 8  & {\bf 1}    & 10  \\
\hline
a=8           & 8  & 9    & 6  & 4   & 10    & 3    & 2  & 5    & 7  & {\bf 1}    & 10  \\
\hline
a=9           & 9  & 4    & 3  & 5 & {\bf 1} & 9    & 4  & 3    & 5  & 1    & 5   \\
\hline
a=10         & 10  & {\bf 1}   & 10  & 1   & 10    & 1   & 10  & 1   & 10  & 1    & 2   \\
\hline
\end{tabular}
\end{center}

@@

Since $45$ is not a prime, there is no ``multiplicative order'' for all values of $a$ (for all numbers that are not 
relatively prime to $45: 3, 5, 6, 9, 10, 12, \cdots,$ because $45 = 3^2*5$).

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\subsection{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``\textbf{RSA}'' $-$ the following remark is meant with humor to show
that each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
%(\hyperlink{appendix-movies}
%{see appendix A3: Movies and Literature with Relation to Cryptography}).
(see appendix \ref{s:appendix-movies}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

@@

We now combine the conjectures from step 1 and 2 for $n=p*q$ to show that:
$$ M^{e * d} \equiv M{\rm ~(mod~}n) {\rm ~for~all~} M < n. $$
From step 1 and 2 we have $(M^{e * d} - M) \equiv 0 {\rm ~(mod~} p)$ and  $(M^{e * d} - M) \equiv 0 {\rm ~(mod~} q)$.
Therefore, $p$ and $q$ are both factors of the same number $z = (M^{e * d} - M)$.
Since $p$ and $q$ are {\bf distinct} prime numbers, their product must also be a factor of this number $z$. Thus:
$$
(M^{e * d} - M) \equiv 0 {\rm ~(mod~}p*q) {\rm ~~or~~ } M^{e * d} \equiv M {\rm ~(mod~}p*q) {\rm ~~or~~} 
 M^{e * d} \equiv M {\rm ~(mod~}n).
$$
\hfill$\Box$

@@

$O(l)$ is the average number of processor steps depending on the bit length
$l$ of the number $n$ to be factorised.  For the best currently known
factorisation algorithm the constant $c = (64/9)^{1/173} = 1923$.

@@

The task of the researchers was not initiated by a challenge, but
they wanted to find the last prime factors of the Mersenne number $2^{1039}+1$
(see ``Wanted List'' of the Cunningham Project\index{%
Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}\\
Cunningham table: \href{http://homes.cerias.purdue.edu/~ssw/cun/pmain1206}
                {\texttt{http://homes.cerias.purdue.edu/\~{}ssw/cun/pmain1206}}\\
The numbers in the Cunningham table have the following syntax:\\
``(2,n)-'' means $2^{n}-1$;~~~
``(2,n)+'' means $2^{n}+1$.\\
To describe the magnitude one writes $p<n>$ or $c<n>$: 
``n'' is the number of decimal digits and ``p'' and ``c'' tell,
whether the number is prime or composite.\\
$2^{1039}-1 = p7 * c307 = p7 * p80 * p227$ \\
It is explained more precisely at the CUN page:\\
``2651+ means $2^{651} + 1$ and the size (c209 means 209 decimal digits)
of the
number which was factored.  Then come the new factor(s), the discoverer and
the method used.  Recently, only the multiple polynomial quadratic sieve
(ppmpqs), the elliptic curve method (ecm) and the number field sieve (nfs)
have been used.  `hmpqs' stands for hypercube multiple polynomial quadratic
sieve.  Under `new factors', `p90' means a 90-digit prime and `c201' is a
201-digit composite number.''.
}).\\

@@

{\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{Appendix_A}{appendix A of this chapter}}. \\

@@

This $28$-digit number is divided into $4$-digit parts
(because $2626$ is still smaller than $n=3713$):\\
{\tt 0120 2001 0311 0001 2000 0401 2314}

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code to do RSA
encryption using Mathematica and Pari-GP. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

{\bf 5. Decryption}: \\
Cipher text: {\tt 1404 2932 3536 0001 3284 2280 2235 }\\
This $28$-digit number is divided into $4$-digit parts.\\
All 7 parts are decrypted using:  $M \equiv C^{97}{\rm ~(mod~}3713)$: \\
{\tt 0120 2001 0311 0001 2000 0401 2314} \\
The 2-digit numbers are transformed into capital letters and blanks.

Using the selected values it is easy for a cryptanalyst\index{Cryptanalysis}
to derive the secret values from the public 
parameters $n=3713$ and $e=37$ by revealing that $3713 = 47 * 79$.\\  
If $n$ is a $768$-bit number, there is, according to present knowledge, little chance of this.

@@

Blocks are formed\footnote{\vskip +3 pt \tt \begin{tabular}{ll@{ }l@{ }l}
& binary representation  & decimal representation\\
01010010, 82 & 01010010 01010011 & = &21075 \\
01010011, 83 & \\
01000001, 65 & 01000001 00100000 & = &16672  \\
00100000, 32  \\
01110111, 119 & 01110111 01101111 & = &30575 \\
01101111, 111 \\ 
01110010, 114 & 01110010 01101011 & = &29291 \\
01101011, 107 \\
01110011, 115 & 01110011 00100001 & = &29473 \\
00100001, 33: 
\end{tabular}} (each ASCII character is encoded into a 8 digit binary number below):\\
{\tt 21075 16672 30575 29291 29473}\footnote{Using CrypTool\index{CrypTool} 
you can solve this with the menu
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration} 
with the following options: all 256 ASCII characters, b-adic, block length\index{Block length} 2 and decimal representation.}

@@

Each block is decrypted using: $M \equiv C^{2473}{\rm ~(mod~}67,519)$: \\
{\tt 82083 65032 119111 114107 115033}
